Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-09-30 demystify.tex
updates

If you pass this, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
 Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot  (because the student often has little experience in writing LoRs) and  \red{red flag} such letters.

Most well-known and well-respected profs would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). This might be a common practice at their university or the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer, relatively unknown in the field, or simply do not know/care enough about you.  In any of these cases, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

\subsection{Letters from Admin people}\label{sec:admin-letters}

Many international students try to get letters from high-ranking admin people in their universities such as department head, dean, or director. The students often mistakenly believe that these letters are more valuable because such people are important in their universities.  However, these people, while well-known and respected in your university, might not be active in research (e.g., they focus on administration) and thus are not recognized by adcom members. In fact, to adcom members, such a letter has little value if the writer does not know you well enough (e.g., did not work with you) to talk about your research ability.

@@

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item [Ivy League] A group of eight private universities in the US known for their academic excellence and social prestige. The Ivy League schools are Brown, Columbia, Cornell, Dartmouth, Harvard, the UPenn, Princeton, and Yale. Note that these schools are typically known for their undergraduate programs more than their graduate programs. Moverover, most top CS programs such as CMU, UIUC, UCSD, UWash, MIT, Stanford, are not in the Ivy League.

  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab is DynaROARS, which focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

@@

  \item[MS] Master of Science. A graduate degree that typically requires 2 years of study.
  \item[Open House] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

  \item[Q1 journal] Commonly mentioned by international students to refer to top-tier journals in CS.  This is not a standard in CS in the US (which often focuses on confs. rather than journals), and many CS faculty in the US might not know what this means.  If you want to refer to top-tier journals, you should mention the specific journals, e.g., TSE in Software Engineering.

@@

\begin{table}
    \caption{PhD vs. MS}\label{tab:phd-vs-ms}
    \begin{tabular}{c|c|c}
    \toprule
    \textbf{} & \textbf{PhD} & \textbf{MS} \\
    \midrule
    Objective & Industry & Research \\
    Admission Req & Research experience & No research experience \\
    Coursework Req & Yes & Yes (research is much more important)  \\
    Duration & 2 years & 5--7 years \\
    Adviser Req & No & Yes \\
    Funding & No & Yes \\
    \bottomrule
    \end{tabular}
  \end{table}

\autoref{tab:phd-vs-ms} summarizes the main differences between PhD and MS programs:
\begin{itemize}

@@

\subsection{Admission Committee}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling (i.e., they review applications as they come in and make decisions throughout the admission cycle).  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does not review much time to review compared to a PhD application (\autoref{sec:ievaluate}).

>>>>

After screening, your application is complete and forwarded to the CS department for further evaluation. If you don't pass screening,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
 Many international students write letters themselves, typically due to the requests of their profs. or letter writers, and have them signed by their profs. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot  (because the student often has little experience in writing LoRs) and  \red{red flag} such letters.

Most well-known and well-respected profs would \emph{not} ask you to write your own letter (in fact, even not well-known ones wouldn't do this). This might be a common practice at specific universities and the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they either do not know/care enough about you or don't know how to write a good letter.  In any case, such letters are not useful and might even hurt your application.  So if you are in this situation, you should find someone else to write for you.

\subsection{Letters from Admin People}\label{sec:admin-letters}

Many international students try to get letters from high-ranking admin people in their universities such as department chair/head, dean, or director. The students often believe that these letters are more valuable because such people are important in their universities.  However, these people, while well-known and respected in your university, might not be active in research (e.g., they focus on administration) and thus are not recognized by adcom members. Such a letter has little value if the writer does not know you well enough (e.g., did not work with you) to talk about your research ability.

@@

  \item[Adcom \autoref{sec:evalapps}] Admission Committee. The group of faculty members who review applications and make admission decisions.

  \item [Adcom members \autoref{sec:evalapps}] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.

  \item[Adcom chair \autoref{sec:evalapps}] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).

  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.

  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.

  \item [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions. Note that you do not need to be from an underrepresented group to write a diversity statement.

  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.

  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.

  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.

  \item [Ivy League] A group of eight private universities in the US known for their academic excellence and social prestige. The Ivy League schools are Brown, Columbia, Cornell, Dartmouth, Harvard, the UPenn, Princeton, and Yale. These schools are typically known for their undergraduate programs more than their graduate programs. Moreover, most top CS programs such as CMU, UIUC, UCSD, UWash, MIT, Stanford, are not in the Ivy League. See selecting and ranking schools in \autoref{sec:selecting-ranking-schools}.

  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab \href{https://dynaroars.cs.gmu.edu}{DynaROARS} focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

@@

  \item[MS \autoref{sec:msrequirement}] Master of Science. A graduate degree that typically requires 2 years of study.

  \item[Open House \autoref{sec:accepted}] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

  \item[Q1 Journal] Commonly mentioned by international students to refer to top-tier journals in CS.  This is not a standard in CS in the US (which often focuses on confs. rather than journals), and many CS faculty in the US might not know what this means.  If you want to refer to top-tier journals, you should mention the specific journals, e.g., TSE in Software Engineering.

@@

\begin{table}
    \caption{MS vs. PhD}\label{tab:phd-vs-ms}
    \begin{tabular}{c|c|c}
    \toprule
    \textbf{} & \textbf{MS} & \textbf{PhD} \\
    \midrule
    Objective & Industry & Research \\
    Admission Req & Research experience & No research experience \\
    Coursework Req & Yes & Yes (research is much more important)  \\
    Duration & 2 years & 5--7 years \\
    Adviser Req & No & Yes \\
    Funding & No & Yes \\
    \bottomrule
    \end{tabular}
  \end{table}

\autoref{tab:phd-vs-ms} summarizes the main differences between MS and PhD programs:
\begin{itemize}

@@

\subsection{Admission Committee}\label{sec:ms-adcom}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling (i.e., they review applications as they come in and make decisions throughout the admission cycle).  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does not review much time to review compared to a PhD application (\autoref{sec:ievaluate}).
